Separating Use and Reuse to Improve Both

### 2017-05-20 FACS/Intro.tex
.

Type systems based on structural typing can deal with the first
situation well, but not the second. Since structural subtyping
accounts for the types of the methods only, a Bag would be a subtype
of a Set if the two interfaces are type compatible. For dealing with
the second situation nominal subtyping is preferable. With nominal
subtyping an explicit subtyping relation must be signalled by the
programmer. Thus if subtyping is not desired, the idea is that 
programmer can simply not declare a subtyping relationship.

While there is no problem in subtyping without subclassing, in the design
of most nominal OO languages subtyping implies subtyping in a
fundamental way. This is because of what we call the self-reference
\emph{leaking problem}. The leaking problem is illustrated by the following
(Java) code:

@@

In class \lstinline{A}, the method \lstinline{m} passes \lstinline{this} as \lstinline{A}.
This code seems correct, and there is no subtyping/subclassing.
Now, lets add a class \Q@B@

@@

%%Class \lstinline{B} does two things at the same time: 
%%1) it {\bf inherits} the method \lstinline{ma} from
%%\lstinline{A}; and 2) it creates a {\bf subtype} of \lstinline{A}.
We can see an invocation of \lstinline{ma} inside
\lstinline{mb}, where the self-reference \lstinline{this} is of type \lstinline{B}. 
The execution will eventually call \lstinline{Util.m} with an
instance of \lstinline{B}. However, this can be correct only if \lstinline{B} is a subtype of
\lstinline{A}.

If Java was to support a mechanism to allow reuse/inheritance 
without introducing subtyping, such as:

\begin{lstlisting}[language=Java]
  class B inherits A{ int mb(){return this.ma();} }
\end{lstlisting}

\noindent Then an invocation of 
\lstinline{mb} would be type-unsafe (i.e. it would 
result in a run-time type error). 
Note that here the intention of using the imaginary keyword {\bf
  inherits} is to allow the code from \lstinline{A} to be inherited 
without \lstinline{B} becoming a subtype of \lstinline{A}. 
However this breaks type-safety. The problem is that the
self-reference \lstinline{this} in class \lstinline{B} has 
type \lstinline{B}. Thus, when \lstinline{this} is passed as an argument to 
the method \lstinline{Utils.m} as a result of the invocation of
\lstinline{mb}, it will have a type that is incompatible with the
expected argument of type \lstinline{A}.  
\bruno{
We can also mention what happens in structural types:
"System with structural types would *be required to* guarantee that
the structural type of A is a supertype of the structural type of B”.
}

What the leaking problem illustrates is that
adopting a more flexible
nominally typed
OO model where subclassing does not imply subtyping is not trivial. 
If we are to separate subclassing from subtyping then a more substantial
change in the language design is necessary.  Every OO language with
the minimal features exposed in the example (using \lstinline{this},
extends and method calls) is forced to accept that subclassing implies
subtyping.\footnote{ Note how this is true also in C++, where is
  possible to "extends privately". Such is a limitation of the
  visibility of subtyping but not over subtyping itself, and the
  former example would be accepted by C++ even if \lstinline{B} was to
  "privately extends" \lstinline{A}.}
In essence we believe that a problem with the design of classes in languages like Java
is that classes do too many things at once. In particular they act
both as units of \emph{use} and \emph{reuse}.
That is, in languages like Java, classes can be
instantiated (using {\bf new}) to create objects that \emph{use} some 
functionality. Classes can also be subclassed to provide \emph{reuse}
of code.

\marco{Cite some work of bruceTHisType and show how he also fails.}

\begin{comment}
An iconic example on why connecting inheritance/code reuse and
subtpying is problematic is provided by the
historic\cite{LaLonde:1991:SSS:110673.110679}: A reasonable
implementation for a \Q@Set@ may be easy to extend into a \Q@Bag@
keeping tracks of how many times an element occurs.  We would just add
some state and override a couple of methods.\bruno{Are we going to
  present this example in the paper solved in 42, for example? 
I think I would expect to see it.}

However, our subclassing would break Liskov substitution principle (LSP)~\cite{martin2000design}: not all bags are sets!
Of course, one could retroactivelly fix this problem by introducing \Q@AbstractSetOrBag@
and making both \Q@Bag@ and \Q@Set@ inherit from it.
This looks unnatural, since \Q@Set@ would extend it without adding anything,
and we would be surprised to find a use of the type \Q@AbstractSetOrBag@.
Worst, if we was to constantly apply this mentalty, we would introduce a very high number
of abstract classes that are not supposed to be used as types, and that will clutter the 
public interface of our classes and our code project as a whole.
\end{comment}

This paper aims at showing a simple language design, called \name, to
completely decouple subtyping and subclassing in a nominally typed OO
language. The key idea is to divide between code designed for
\textbf{USE} and code designed for \textbf{REUSE}. 
In \name there are two separate concepts: classes
and traits~\cite{}. Classes are meant for code use, and cannot be used
for reuse. In some sense classes in \name are like final classes in
Java. Traits are meant for code reuse and multiple traits can be
composed to form a class which can then be instantiated. Traits 
cannot be instantiated (or used) directly. Such design allows the
subtyping and code reuse to be treated separately, which in terms 
brings several benefits in terms of flexibility and code reuse.
\marco{We need to talk of unanticipated extensions?}
\bruno{Talk more about typing aspects here. Summarize the important 
aspects of the design of \name.}

>>>>

Type systems based on structural typing can deal with the first
situation well, but not the second. Since structural subtyping
accounts for the types of the methods only, a Bag would be a subtype
of a Set if the two interfaces are type compatible. For dealing with
the second situation nominal subtyping is preferable. With nominal
subtyping an explicit subtyping relation must be signalled by the
programmer. Thus if subtyping is not desired, the idea is that 
programmer can simply {\bf not} declare a subtyping relationship.

While there is no problem in subtyping without subclassing, in the design
of most nominal OO languages subtyping implies subtyping in a
fundamental technical way. This is because of what we call the
\emph{this leaking problem}, illustrated by the following
(Java) code:

@@

Method \lstinline{A.ma} passes \lstinline{this} as \lstinline{A} to \Q@Util.m@.
This code seems correct, and there is no subtyping/subclassing.
Now, lets add a class \Q@B@

@@

%%Class \lstinline{B} does two things at the same time: 
%%1) it {\bf inherits} the method \lstinline{ma} from
%%\lstinline{A}; and 2) it creates a {\bf subtype} of \lstinline{A}.
We can see an invocation of \lstinline{A.ma} inside
\lstinline{B.mb}, where the self-reference \lstinline{this} is of type \lstinline{B}. 
The execution will eventually call \lstinline{Util.m} with an
instance of \lstinline{B}. However, this can be correct only if \lstinline{B} is a subtype of
\lstinline{A}.

%If Java was to support a mechanism to allow reuse/inheritance 
%without introducing subtyping, such as:
%
%\begin{lstlisting}[language=Java]
%  class B inherits A{ int mb(){return this.ma();} }
%\end{lstlisting}
%
%\noindent Then an invocation of 
%\lstinline{mb} would be type-unsafe (i.e. it would 
%result in a run-time type error). 
%Note that here the intention of using the imaginary keyword {\bf
%  inherits} is to allow the code from \lstinline{A} to be inherited 
%without \lstinline{B} becoming a subtype of \lstinline{A}. 
%However this breaks type-safety. The problem is that the
%self-reference \lstinline{this} in class \lstinline{B} has 
%type \lstinline{B}. Thus, when \lstinline{this} is passed as an argument to 
%the method \lstinline{Utils.m} as a result of the invocation of
%\lstinline{mb}, it will have a type that is incompatible with the
%expected argument of type \lstinline{A}.

If in Java code reuse was not introducing subtyping an invocation of 
\lstinline{B.mb} would result in a run-time type error.
The problem is that the
self-reference \lstinline{this} in class \lstinline{B} has 
type \lstinline{B}. Thus, when \lstinline{this} is passed as an argument to 
the method \lstinline{Utils.m} as a result of the invocation of
\lstinline{mb}, it will have a type that is incompatible with the
expected argument of type \lstinline{A}.

Every OO language with the minimal features exposed in the example (using \lstinline{this},
extends and method calls) is forced to accept that subclassing implies
subtyping.\footnote{ Note how this is true also in C++, where is
  possible to "extends privately". Such is a limitation of the
  visibility of subtyping but not over subtyping itself, and the
  former example would be accepted by C++ even if \lstinline{B} was to
  "privately extends" \lstinline{A}.}

In essence we believe that a problem with the design of classes in languages like Java
is that classes do too many things at once. In particular they act
both as units of \emph{use} and \emph{reuse}.
That is, in languages like Java, class names can be
\emph{use}d as type and instantiated.
Classes can also be subclassed to provide \emph{reuse}
of code.

Adopting a more flexible nominally typed OO model where subclassing does not imply subtyping is not trivial,
and a more substantial change in the language design is necessary.

This paper aims at showing a simple language design, called \name, to
completely decouple subtyping and subclassing in a nominally typed OO
language. The key idea is to divide between code designed for
\textbf{USE} and code designed for \textbf{REUSE}. 
In \name there are two separate concepts: classes
and traits~\cite{}. Classes are meant for code use, and cannot be used
for reuse. In some sense classes in \name are like final classes in
Java. Traits are meant for code reuse and multiple traits can be
composed to form a class which can then be instantiated. Traits 
cannot be instantiated (or used) directly. Such design allows the
subtyping and code reuse to be treated separately, which in turn
brings several benefits in terms of flexibility and code reuse.
%\marco{We need to talk of unanticipated extensions?}
%\bruno{Talk more about typing aspects here. Summarize the important 
%aspects of the design of \name.}

### 2017-05-20 FACS/RelatedWork.tex
new abstract
.

>>>>

Add that: original trait model has no self construction and types introduction not relevant since dynamicly typed language

\bruno{
We can also mention what happens in structural types:
"System with structural types would *be required to* guarantee that
the structural type of A is a supertype of the structural type of B”.
}

\marco{Cite some work of bruceTHisType and show how he also fails to separe reuse and use}
